ESM 505/ESR 605 : Data Management, Spring 2014

Module 4: Filtering and Subsetting Data
====================

We briefly discussed selecting data within an object in Module 2.  Here we will discuss more advanced methods to subsetting data. Because most of the data we will be using in R will be in the form of data frames we will concentrate on data frame methods for filtering and subsetting.


##  Determining the structure of an object

In order to understand how to select portions of a data object, you will need to know details about the structure of that object first.  The str() function accomplishes that.

```{r}
# Create a series of equally sized vectors
sites<-c("Site A", "Site B")
dates<-as.Date(c("12Mar2013","04Apr2013"),"%d%b%Y")
lat=c(45, 47)
lon=c(-124, -125)
treatment=c(T,F)

# Combine the vectors into a data frame
dat<-data.frame(sites, dates, lat, lon, treatment)

# Use str() on some of the vectors and the data frame
str(sites)
str(lat)
str(dat)

# The name() function is similar but only provides the names of the various data frame columns
names(dat)


# You can also use name() to reassign the names of the data frame columns

names(dat)<-c("SiteName", "VisitDate", "Latitude", "Longitude", "TreatmentApplied")

str(dat)

```

##  Selecting Variables (data frame columns)

### Selecting Single columns

```{r}

# By index
dat[1]

# By Name
dat["SiteName"]

# Using the $ operator

dat$SiteName


```

### By Name

``` {r}

dat[c("SiteName","TreatmentApplied")]

# Alternatively you can use matrix R,C notation

dat[,c("SiteName","TreatmentApplied")]

```

### By column number like a matrix

```{r}

dat[,c(1,5)]

```

### By Logical Assignment

```{r}
dat[c(T,F,F,F,T)]

```

##  Droping Variables

### Excluding Variables

```{r}

# we can use negative indexes

dat[c(-1,-4)]

# or we can use logical assignment

dropcols<-names(dat) %in% c("Latitude","Longitude")
dropcols

dropcols2<-c("Latitude","Longitude")

# this is the similar to a logical assignment for selecting columns but instead a "True" is a column we want to drop 
dat[c(F,F,T,T,F)]

# We use the ! to denote a logical negation or "not"
# Here we specify that we want to select "not the dropcols"
dat[!dropcols]

```


### Dropping Variables permentantly

```{r}

# One option is reassignment

dat<-dat[!dropcols]

# rebuild the data frame
dat<-data.frame(sites, dates, lat, lon, treatment)
names(dat)<-c("SiteName", "VisitDate", "Latitude", "Longitude", "TreatmentApplied")

# the other option is to set some columns to null

dat$TreatmentApplied <-NULL
dat
# But you can't do this with more than one column at a time

dat[c("Latitude","Longitude")]<-NULL

# However you can chain these

dat$Latitude<-dat$Longitude<-NULL
dat


```

##  Selecting Observations (data frame rows)

``` {r}
# loading some real data with lots of records

data2<-read.csv("data/wk1samples.csv")

str(data2)

```

### By index

```{r}

# select first two records
data2[1:2,]

# selecting second and fourth records
data2[c(2,4),]

# selecting not the third and fifth record
data2[c(-3,-5),]

```

### By value

``` {r}

# Selecting data2 where there weren't any fish
data2[data2$fish == 0,]

# Using & to combine conditions
data2[data2$fish > 9 & data2$siteID == 2,]

```

##  Selecting Observations and Variables

``` {r}
# Sites and dates where we collected zero fish

data2[data2$fish == 0,c("siteID","date")]

# sample IDs where zero fish were collected
data2[data2$fish == 0,]$sampleID

```

##  Subset Function

The subset() function is a shortcut to selecting observations and variables

``` {r}
# this isn't any shorter than the previous method of finding the sites and dates where we collected zero fish, but it's a bit more readable.

subset(data2, fish == 0, select=c(siteID,date))

# we can also drop columns using the select option

subset(data2, fish==0, select=c(-siteID, -date))

# Subset makes it easier to combine multiple conditions

subset(data2, fish==0 & siteID ==1 & date == "2008-03-01")

# You can even do "OR" conditions using the "|" operator
# here we are selecting where fish collection was less than or equal to 3 or greater than 11

subset(data2, fish <=3 | fish > 11)




```
 Remember you can combine a series of logical operators ( "|" and "&")
 it helps if you know some logic to help simply your expressions
 for example...
 
 !(a & b) is the same as !a | !b
 !(a | b) is the same as !a & !b 

```{r}

 subset(data2, !(fish  == 3 & siteID == 1))
 
 # is the same as 

 subset(data2, !fish == 3 | !siteID==1)

 # and

 subset(data2, !(fish == 3 | siteID == 2))

 # is the same as 

 subset(data2, !fish == 3 & !siteID == 2)



``` 

You can also use the %in% operator to match multiple values at once.

```{r}

dates_of_interest<-c("2008-02-22", "2008-03-01")
subset(data2, date %in% dates_of_interest)
 
# is the same as 
subset(data2, date == "2008-02-22" | date == "2008-03-01")


```



##  Random samples
Sometimes you will want a random subset of your data.  To do this, use the sample() function without replacement.

```{r}

# first determine how many rows of data you have
rowcount<-nrow(data2)

# select two random rows
randomrows<-sample(rowcount, 2)

# use the "randomrows" to select those rows from the data frame
data2[randomrows,]


``` 

## Head and Tail

Large data frames take a long time to print out to the screen.  Because of this, sometimes you just want to view the first portion (the head) or the last portion (the tail) of a data frame.  To get a sense of what the data looks like.   

```{r}

# load a rather large dataset
load("data/WSA_data.Rdata")

# How many rows of data using nrow()
nrow(data)

# To make my tables neat lets select just the 5 columns of data
data<-data[,1:5]

# The head (beginning) of the data
head(data)


# The tail (end) of the data
tail(data)

# the defualt is 6 rows of data, but you can change this by specifing an n value

head(data, n=20)


```
##  Assigning and Subsetting

Sometimes you want to select data so that you can reassign values in the data frame. This is useful when you are cleaning up your data.

```{r}

# for example, sometimes you want to replace NA values with 0's

data3<-data.frame(site=c(1,1,1,2,2,2),count=c(2,4,3,NA,5,NA))
data3

# to select rows where count is NA you can use the is.na() function

data3[is.na(data3$count),]

# to set count to zero where count was NA

data3[is.na(data3$count),]$count<-0

data3

```

##  Ordering

While not exactly a form of subsetting, ordering of data frames is a related action.  We can use the order() and sort() functions to accomplish this.

```{r}

# A simple ordering of a vector
x<-c(2,3,5,3,2,1)
order(x)

# Well that's now what we want. What is happening here?  The order function returns the values of the vector index in order of the values we specified.  In this case the 6th value was the lowest (first) and the 3rd value was the highest (last).  To get the results we wanted we need to use order in a slightly different way.

x[order(x)]


# Using sort() gives us what we probably were expecting.

sort(x)



#  Then why bother with order().  Well, sort() doesn't work on data frames.   

data3

# to sort the data3 data frame by count

data3[order(data3$count),]


```


## Bonus Material

The following PDF is a 6 page cheat sheet for many of the basic operators, functions, and other important bits of R.  I try to keep a copy handy when I'm coding in R in case I need a reminder on which function I might need for a particular job.

http://cran.r-project.org/doc/contrib/Baggott-refcard-v2.pdf


## Homework

Using the data set you found for Module 2 and imported in Module 3 write a script to create 3 interesting subsets of data that you might want to use for later analysis.  Use which ever methods make the most sense to you but be sure to assign these subsets to new variables with meaningful names. Remember to add code from Module 3 regarding how to import this data into the beginning of your new script.

1.) Upload your data and script to this module's dropbox on d2l.




