Reproducible Research, Spring 2021

Module 3b : Dealing with Dates, Strings, and Factors
=======================


## Lubridate

### Dealing with dates

Dealing with dates in base are is a hassle.  Dates often get imported as readable strings that mean nothing to the computer. 
```{r}
string_date <- "28 August 1972"
string_date2 <- "09/22/2002"


```
Alternatively, dates exist that are readable by the computer but are ugly to look at.

```{r}
mynow <- Sys.time()
mynow
str(mynow)

```
What's a POSIXct format?  It's one of many date/time formats in R and it's annoying.
In base R, we can output that into something readable using format.date(), but you have to remember a bunch of formatting codes.

```{r}
format.Date(mynow, "%m/%d/%Y")


```
Likewise those format codes can be used to turn a string date into a Date
```{r}

str(string_date)
mydate <- as.Date(string_date, "%d %B %Y")
str(mydate)
```

There are a bunch of base R functions to deal with formatting and manipulating dates and times, but the lubridate package simplifies that.

Here's how to convert string dates into the date data type.
```{r}
library(lubridate)
mydate <- dmy(string_date) 
mydate
str(mydate)
mydate2 <- mdy(string_date2)
mydate2
str(mydate2)

```
Lubridate is very robust and can handle variation that often creeps into data.
```{r}
mydates <- c("10/22/2003", "Oct 23, 2004", "11-2-2005")
mydates <- mdy(mydates)
mydates

```


For output, Lubridate allows you to use a human friendly template called stamp()
```{r, message=FALSE}
date_stamp <- stamp("Sunday, March 1, 2021")
date_stamp(mydates)


```

### Date Math

Lubridate makes math with dates easy.

```{r}

# difference between two dates
mydates - ymd("2000-03-01")

# subtract days or weeks from a date.  A date is returned
mydates - days(35)
mydates - weeks(3)

# subtract an array of days from a date and an array of dates is returned
mydates[1] - days(1:5)

# can do date math without a specified date, just units of time
out <- months(3) - weeks(2) + days(3) - hours(6)
out
as.period(out, "day")
as.period(out, "minute")

```



Here's a cheat sheet for lubridate
https://evoldyn.gitlab.io/evomics-2018/ref-sheets/R_lubridate.pdf



## Stingr

### Dealing with strings and characters
```{r}
library(stringr)

# create an array of strings
x <- c("Ty", "Tom", "John","Philip","Alexander")

# how long is each string
str_length(x)

# collpase the array of strings into a single string
str_c(x, collapse = ", ")
str_c(x, collapse = "####")

# get the 1-2 character of each string
str_sub(x, 1, 2)

# select the strings in the array that have an x,y, or z
str_subset(x, "[xyz]")

# do the strings contain P or J
str_detect(x, "[PJ]")

# how many vowels does each string have
str_count(x, "[aeiou]")

# replace the first vowel of each string with a '#'
str_replace(x, "[aeiou]","#" )

# replace all the vowels in each string with '#'s
str_replace_all(x, "[aeiou]", "#")

# pad the strings so that that are all 10 characters long, check the length
str_pad(x, 10, "right")
str_length(str_pad(x, 10, "right"))

# truncate any string over 7 characters long, eding with '...'
str_trunc(x, 7)

# you can specify the ellipsis character
str_trunc(x, 3, ellipsis = '#')


# replacing, trimming, and squishing strings
z = " hello    there "
str_replace_all(z, " ", "")
str_trim(z)
str_squish(z)


# changing letter case
b <- "oNCe UPon A timE."
str_to_lower(b)
str_to_upper(b)
str_to_title(b)
str_to_sentence(b)


# Using str_glue() instead of paste()
species <- "Carcinus maenas"
count <- 120
catch_date <- ymd("2020-10-12")

paste("We caught", count, species, Sys.Date() - catch_date, "days ago!", sep=" ")

str_glue("We caught {count} {species} {Sys.Date() - catch_date} days ago!")



```

More can be done if you learn a bit about regular expressions.   


http://edrub.in/CheatSheets/cheatSheetStringr.pdf


## Forcats

### Dealing with factors (categorical variables)

Factors generally aren't a problem until you try to put them in a table or graph.
In particular, it's often nice to have bar plots ordered from low to high.  We'll revist this often when we start plotting with ggplot2.

fct_reorder(): Reordering a factor by another variable.
fct_infreq(): Reordering a factor by the frequency of values.
fct_relevel(): Changing the order of a factor by hand.
fct_lump(): Collapsing the least/most frequent values of a factor into “other”.


```{r}
library(forcats)
library(dplyr)
starwars %>% count(species)
# That was not in any order.  Try fct_infreq() to order by frequency
starwars %>% mutate(species = fct_infreq(species)) %>% count(species)
# better order, still too many species, let's lump any species with 2 or less
starwars %>% filter(!is.na(species)) %>% mutate(species = fct_infreq(species)) %>% mutate(species = fct_lump_min(species, 2)) %>% count(species)
# better order and lumped all infrequent species together as "other"
```


```{r}
starwars %>% mutate(height_bin = cut(height, breaks = c(0,100,200,300), labels=c("small","medium","large"))) %>% count(height_bin)

```

## Homework

Explore the dates, strings, and factors in your data.     Upload your R script to the d2l dropbox for this module.


