ESM 505/ESR 605 : Data Management, Spring 2014

Week 3: Creating, Importing, & Exporting Data
====================

## 1 Creating Data

### 1.1 Vectors

To create a vector of values in R you use the concatenate "c" function.

```{r}
#numeric  vector
nv<-c(2,3,4,2)
print(nv)

#character vector
cv<-c("put","some","words","here")
print(cv)

#logical vector
lv<-c(T,F,F,T)
print(lv)
```

### 1.2 Matrices
The main methods to create matrices are the "matrix" function and the use of the "cbind" or "rbind" functions.

The matrix function takes a vector (creatd with the "c" function) and then provides information on how many rows and/or columns to divide it into.  

```{r}
mat<-matrix(c(1,2,3,4,5,6,7,8,9), nrow=3)
print(mat)

```
If you supply a number or rows that doesn't evenly divide the elements of the vector and error will appear. 

```{r}
matrix(c(1,2,3,4,5,6,7,8,9), nrow=2)

```

Using cbind or rbind to bind columns or rows will also work.  If the vectors you are binding are of the same class (numeric, character, logical), the resulting matrix will also be that class.  However, if the vectors differ the matrix will defualt to the most compatible type.  

```{r}

mat2<-cbind(nv,cv,lv)
print(mat2)
mode(mat2)

mat3<-cbind(nv,lv)
print(mat3)
mode(mat3)

```


### 1.3 Data Frames
Most imported data will end up a data frame by default.  However, you can create a data frame from other R objects using the "data.frame" function.  When making a data frame, the columns need to all be the same length and the rows need to be the same length.

```{r}

names<-c("Fred","Bob","Bill","Jim")
weight<-c(129,145,234,198)
height<-c(64,68,72,70)

data<-data.frame(names,weight,height)
print(data)

```



### 1.4 Lists


```{r}
data2<-list(name,weight,height)
print(data2)

data3<-list(name, weight, height, data)
```

### 1.5 Converting between types

```{r}
a<-c("10","20","30")
print(a)
class(a)

b<-as.numeric(a)
print(b)
class(b)

c<-as.factor(b)
print(c)
class(c)

```




## 2 Importing Data

Getting your data into R can sometimes be the hardest part.  Luckily there are packages and functions to help with this.

### 2.1 Copy/Paste

If you try to copy and paste some data in from a text file or an Excel spreadsheet and assign it to an object you just end up with a character vector with a single element made up of long string of characters. 

``` {r}
HaresLynx <- "Year  Hares  Lynx
1900 30 4
1901 47.2 6.1
1902 70.2 9.8
1903 77.4 35.2
1904 36.3 59.4
1905 20.6 41.7"

print(HaresLynx)

```
Probably not what you wanted.

To remedy this we use read.table and TextConnection to parse the characters into a data frame

```{r}
HaresLynx2 <- read.table(textConnection(HaresLynx), header=TRUE)
print(HaresLynx2)

```


### 2.2 Text Files
Probably the easiest way to import data is via text file.  I find CSV files to work the best..

```{r}
sites<-read.csv("data/wk1sites.csv")
print(sites)

```



### 2.3 Database Connection

Connections can be made directly with a database.  However, which databases you can connect to is dependent on your operating system.  For example, you need to be using Windows to connect to a MS Access database.  

One common database is MySQL.  We can access MySQL databases using either the "RODBC" library or the "RMySQL" library 

```{r}

library(RMySQL)
# make a connection to the database
con<-dbConnect(MySQL(), user="root", password="", dbname="nemesis_development", host="localhost")

# list the tables in that database
dbListTables(con)

# run a select query on a table in the database and limit it to 10 records
taxa<-dbGetQuery(con, "select id, binomial, taxa_group from taxa limit 10;")
print(taxa)

```


### 2.4 Other formats (JSON, XML, etc..)

Sometimes you'll find data in other formats.  It's worth looking to see if a package has been made to help import that kind of data.

## 3 Exporting Data

### 3.1 Text Files
An analog to read.csv is write.csv.  You can write just about any data frame into a csv file with this command

```{r}

write.csv(taxa, "taxa.csv")

```


### 3.2 Database Connection

You can run append, insert, and update queries using a connection to the database.   You can also write a data frame into a table useing "dbWriteTable(con, "table name", a.data.frame)

### 3.3 Other formats (JSON, XML, etc..)

The same specialized R packages you used to import other types of data will likely have a function for writing that type of data from R objects.


### 3.4 dput function

It is often useful to useful to export an R object into an R statement that can be used to recreate that object elsewhere.  For example, say you wanted to ask an R question on http://stackoverflow.com/ and needed to include some small snippit of your data to help explain your issue.   Just doing a copynig and pasting will cause formatting issues and it will be difficult for others to usee your data to help you out.   However, if we use the "dput" function we can transform an R object into a the command we'd need to recreate that structure.

```{r}
dput(taxa)

#If I copy the output to that and assign it to an object..

taxanew<-structure(list(id = c(-578L, -577L, -576L, -575L, -574L, -572L, 
-571L, -570L, -569L, -568L), binomial = c("Dasya sp. A", "Dasya sessilis", 
"Pkea yoshizakii", "Chondracanthus teedei", NA, "Tricellaria inopinata", 
"Pachycordyle michaeli", "Gambusia holbrooki", "Corella inflata", 
NA), taxa_group = c("Crustaceans-Copepods", "Algae", "Algae", 
"Algae", NA, "Ectoprocts", "Coelenterates-Hydrozoans", "Fishes", 
"Tunicates", NA)), .Names = c("id", "binomial", "taxa_group"), row.names = c(NA, 
10L), class = "data.frame")


# I get a copy of the original object
print(taxanew)

```

## Homework

1.   Import the dataset you found for you last homework into R. 

2.   Submit both the your found datast and an R script of the code you used to import it onto d2l


